Last glacial cycle glacier erosion potential in the Alps

### 2021-01-24 articles/erosion/alps-erosion.tex
Rework methods and results for new erosion law.

    The ice-sheet model set-up was presented, and the results discussed, in an
    earlier publication \citep{Seguinot.etal.2018}. The simulations use the
    Parallel Ice Sheet Model (PISM, development version~e9d2d1f), an open
    source, finite difference, shallow ice sheet model
    \citep{PISM-authors.2017}. The model includes temperature and water-content
    dependent creep, pseudo-plastic basal sliding accounting for till
    dilatation under high water pressure, bedrock deformation under the ice
    load, and a~positive degree-day (PDD) surface mass balance model. The model
    is initialized with assumed present-day ice thickness and equilibrium
    ice and bedrock temperature at 120\,ka, and ran to the present.
    Climate forcing is provided by a~monthly climatology from interpolated
    observational data \citep[WorldClim;][]{Hijmans.etal.2005} and the European
    Centre for Medium-Range Weather Forecasts Reanalysis Interim
    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature lapse-rate
    corrections from the European Project for Ice Coring in Antarctica
    \citep[EPICA;][] {Jouzel.etal.2007}, and in some cases, time-dependent
    palaeo-precipitation reductions. Lower-resolution, alternative climate
    scenarios are analyzed in the discussion part.

@@

    Modelled erosion rates, $\dot{e}$, are calculated from the modelled basal
    velocities, $u_\mathrm{b}$, using a near-quadratic empirical erosion law,
    %
    \begin{equation}
        \dot{e} = K_\mathrm{g} u_\mathrm{b}^l ,
    \end{equation}
    %
    where $K_g = 2.7\times 10^{-7}\,m^{1-l}\,a^{l-1}$ and $l = 2.02$ are
    empirical constants fitted to measured suspended sediment loads and
    reconstructed basal velocities of Franz-Joseph Glacier
    \citep{Herman.etal.2015}. It is important to note that there are both large
    uncertainties around these numbers concerning Franz-Joseph Glacier
    itself \todo{quantify}, and little insurance that this erosion law applies
    elsewhere.

@@

    The modelled cumulative (time-integrated) glacial erosion potential
    (Fig.~\ref{fig:cumulative}a) varies by several orders of magnitude
    from insignificant to (unrealistic, see below) kilometre-scale erosion
    potential. Its spatial patterns of modelled erosion potential show a very
    strong localization along the Alpine valleys, with local maxima at
    the Alpine gates where ice flow transited from valley to piedmont glaciers,
    and sometimes further up-valley.
    There is a genereral tendency for higher cumulative erosion in the
    north-western Alps where the input winter precipitation
    \citep[WorldClim,][Fig.~1h]{Seguinot.etal.2018} is higher and the
    preexisting glacial topography more pronounced.

@@

    The modelled volumic (domain-integrated) glacial erosion rates
    (Fig.~\ref{fig:cumulative}b) do not correlate with the modelled total ice
    volume. A comparison between the modelled ice volume and the modelled
    volumic erosion rate show no relation between these two quantities, except
    for the onset and termination of the glacial cycle were ice volume is low
    and sliding processes unlikely to be captured by the model resolution.
    Instead, erosion is modelled to be higher by a factor 3 to 10 during
    periods of glacier retreat, than during periods of glacier advance
    (Fig.~\ref{fig:evolution}).

@@

    \note{I suddenly realize it may be better to highlight results from the
          Koppes et al. erosion law for tidewater glaciers instead, and maybe
          pull the sensitivity study up before the main results. To dedice
          about this I will need to get back to the Koppes and Cook paper to
          really understand what these laws are about.}

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Sensitivity to climate}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Erosion uncertainties}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

>>>>

    The ice-sheet model set-up was presented, and the results discussed, in an
    earlier publication \citep{Seguinot.etal.2018}. The simulations use the
    Parallel Ice Sheet Model (PISM, development version~e9d2d1f), an open
    source, finite difference, shallow ice sheet model
    \citep{PISM-authors.2017}. The model includes temperature and water-content
    dependent creep, pseudo-plastic basal sliding accounting for till
    dilatation under high water pressure, bedrock deformation under the ice
    load, and a~positive degree-day (PDD) surface mass balance model. The model
    is initialized with assumed present-day ice thickness and equilibrium
    ice and bedrock temperature at 120\,ka, and ran to the present.
    Climate forcing is provided by a~monthly climatology from interpolated
    observational data \citep[WorldClim;][]{Hijmans.etal.2005} and the European
    Centre for Medium-Range Weather Forecasts Reanalysis Interim
    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature lapse-rate
    corrections from the European Project for Ice Coring in Antarctica
    \citep[EPICA;][] {Jouzel.etal.2007}, and in some cases, time-dependent
    palaeo-precipitation reductions. Lower-resolution, alternative climate
    scenarios are analyzed in the discussion part
    (Sect.~\ref{sec:sensitivity}).

@@

    Modelled erosion rates, $\dot{e}$, are calculated from the modelled basal
    velocities, $u_\mathrm{b}$, using a empirical erosion power-law,
    %
    \begin{equation}
        \dot{e} = K_\mathrm{g} u_\mathrm{b}^l ,
    \end{equation}
    %
    where $K_g = 5.2\times 10^{-11}\,m^{1-l}\,a^{l-1}$ and $l = 2.34$ are
    empirical constants calibrated on quantified sediment yields and velocities
    from 15~Patagonian and Antarctic Peninsula tidewater glaciers
    \citep{Koppes.etal.2015}. It is important to note that there are both large
    uncertainties surrounding these numbers on the aforementioned glaciers
    \todo{quantify}, and little insurance that this erosion law applies
    elsewhere. Alternative erosion laws are analyzed in the discussion part
    (Sect.~\ref{sec:uncertainty}).

    \note{I initially built the paper around the erosion law by
          \citep{Herman.etal.2015}, that yielded similar erosion patterns but
          up to several km of cumulative erosion in the valleys.}

@@

    The modelled cumulative (time-integrated) glacial erosion potential
    (Fig.~\ref{fig:cumulative}a) varies by several orders of magnitude
    from insignificant to hundred-metres scale erosion
    potential. Its spatial patterns show a very
    strong localization along the Alpine valleys, with local maxima occuring both at
    the Alpine gates where ice flow transited from valley to piedmont glaciers,
    and further up-valley where valley slopes increase.
    There is a genereral tendency for higher cumulative erosion in the
    north-western Alps where the input winter precipitation
    \citep[WorldClim,][Fig.~1h]{Seguinot.etal.2018} is higher and the
    preexisting glacial topography more pronounced.

@@

    The modelled volumic (domain-integrated) glacial erosion rates
    (Fig.~\ref{fig:cumulative}b) do not correlate with the modelled total ice
    volume. A comparison between the modelled ice volume and the modelled
    volumic erosion rate shows no single relation between these two quantities, except
    for the onset and termination of the glacial cycle were ice volume is low
    and sliding processes unlikely to be captured by the model resolution.
    Instead, erosion is modelled to be higher by a factor 3 to 10 during
    periods of glacier retreat, than during periods of glacier advance
    (Fig.~\ref{fig:evolution}).

@@

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Sensitivity to climate}
\label{sec:sensitivity}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Erosion uncertainties}
\label{sec:uncertainty}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
